# Paired Two-Sample Tests

Data for paired two-sample tests have the following properties:

- each subject $i$ has two observations $X_i$ and $Y_i$ to give a paired observation $(X_i, Y_i)$
- $(X_1, Y_1), \dots, (X_n, Y_n)$ are independent

Our general strategy is to perform one-sample tests on the paired difference $Z_i = Y_i - X_i$

For this section, we will use the R dataset `sleep`.
In this data, each subject was given both drug 1 and drug 2,
and the change in their sleep was measured.
The column `group` indicates which drug the subject took,
and the column `extra` indicates the change in sleep.

```{r}
data("sleep")

# X: effect on sleep from drug 1
x <- sleep$extra[sleep$group == 1]

# Y: effect on sleep from drug 2
y <- sleep$extra[sleep$group == 2]

# Z: paired difference between X and Y
z <- y - x
```

For each of these tests, we will be testing the hypotheses
$H_0$: $\mu = 0$ vs $H_a$: $\mu \neq 0$,
where $\mu$ is the mean of $Z$.

## Paired $t$-test

- Parametric test
- One-sample $t$-test on the mean $\mu$ of $Z$
- Assumption: $Z_i \overset{\mathrm{iid}} \sim N(\mu, \sigma^2)$
- Hypotheses: $H_0$: $\mu = \mu_0$ vs $H_a$: $\mu \neq \mu_0$ (or $<$, $>$)
- Test statistic: $$T_{\mathrm{obs}} = \frac{\bar{Z} - \mu_0}{s_Z / \sqrt{n}},$$
  where $\bar{Z}$ and $s_Z$ are the sample mean and sample standard deviation
  of $Z$
- Null distribution: Under $H_0$, $T_{\mathrm{obs}} \sim t_{n-1}$

There are three ways to do this test in R using the function `t.test`.

If the observations for group $X$ and group $Y$ are stored in separate vectors `x` and `y`,
then we use `t.test` with the argument `paired = TRUE`.

```{r}
t.test(x, y, paired = TRUE, alternative = "two.sided")
```

If the paired differences $Z = Y - X$ are stored in one vector
`z`, then we use a one-sample `t.test`.

```{r}
t.test(z, alternative = "two.sided")
```


If the data is stored in one matrix or dataframe
where one column contains the observations and another column contains the groups,
we can use the following syntax.
For this data, `extra` contains the observations, and `group` indicates
which group each observation comes from.

```{r}
t.test(extra ~ group, paired = TRUE, data = sleep, alternative = "two.sided")
```

The result for all three values is the same:
$p\text{-value} = 0.002833$.

## Sign test

- Also called binomial test
- Nonparametric test on the median $\theta$ of $Z$
- Assumption: $Z_i$ are iid
- Hypotheses: $H_0$: $\theta = \theta_0$ vs $H_a$: $\theta \neq \theta_0$ (or $<$, $>$)
- Test statistic: $$X = I\{Z_i > \theta_0\},$$
  where $I\{Z_i > \theta_0\}$ is the number of observations $Z_i$ that are greater than $\theta_0$.
- Null distribution: Under $H_0$, $X \sim \mathrm{Binom}(n, p = 0.5)$

The alternative hypothesis $H_a$: $\theta < 0$
is equivalent to $H_a$: $p < 0.5$, where $p = P(Z_i > 0)$.
This is because, under $H_0$: $\theta = 0$, the probability of $Z_i$ being less than the
median $\theta$ is 0.5.

The test is conducted as follows.
```{r}
# calculate paired differences
z <- y - x

# calculate the number of observations greater than 0
X <- sum(z > 0)

# calculate p-value
2 * min(pbinom(X, size = length(z), prob = 0.5),
        1 - pbinom(X, size = length(z), prob = 0.5))
```

## Wilcoxon Signed-Rank Test

- Nonparametric test on the median $\theta$ of $Z$
- Assumptions: $Z_i$ are iid and symmetric
- Hypotheses: $H_0$: $\theta = \theta_0$ vs $H_a$: $\theta \neq \theta_0$ (or $<$, $>$)
- Test statistic: $$V = \sum_{i = 1}^n I\{Z_i > \theta_0\} \mathrm{rank}(|Z_i - \theta_0|)$$

Like with the paired $t$-test, there are three ways to conduct the Wilcoxon test
using `wilcox.test`.

```{r}
wilcox.test(x, y, paired = TRUE, alternative = "two.sided")

wilcox.test(z, alternative = "two.sided")

wilcox.test(extra ~ group, paired = TRUE, data = sleep, alternative = "two.sided")
```

Using all three methods, $p\text{-value} = 0.009091$.

## Permutation Test

- Paired data $(X_i, Y_i)$, where $X_i$ and $Y_i$ are correlated
- Assumption: $(X_1, Y_1), \dots, (X_n, Y_n)$ are iid
- For $n$ subjects, there are $R = 2^n$ equally likely permutation outcomes
- For each possible outcome $\ell \in \{1, \dots, R\}$, we calculate the sample mean of differences
  $$z_\ell = \frac{1}{n} \sum_{i=1}^n (Y_{\ell i} - X_{\ell i}),$$
  where $(X_{\ell i}, Y_{\ell i})$ is the $i$th observation of the $\ell$th permutation outcome.

The $p$-value depends on the direction of $H_a$.

$H_a$        $p$-value
------------ ----------------------------------
$\mu > 0$    $I\{z_\ell \geq \bar{Z}\} / R$
$\mu < 0$    $I\{z_\ell \leq \bar{Z}\} / R$
$\mu \neq 0$ $I\{|z_\ell| \geq |\bar{Z}|\} / R$
------------ ----------------------------------

### Exact permutation test

The exact permutation test for $H_a$: $\mu \neq 0$ is performed as follows.

```{r}
# paired differences
z <- y - x
# sample size
n <- length(z)

# create permutation matrix
lists <- split(matrix(c(-1, 1), n, 2, byrow=TRUE), 1:n)
all.outcomes <- as.matrix(expand.grid(lists))

# observed test statistic
zbar <- mean(z)

# test statistic for permutation samples
zl <- apply(all.outcomes, 1, function(u) mean(u*abs(z)))

# p-value
mean(abs(zl) >= abs(zbar))
```

In this example, we used the sample mean as our test statistic,
but the choice of test statistic for the permutation test is flexible.

### Large-sample approximation for permutation test

Instead of using all $R = 2^n$ possible permutation outcomes,
we can randomly select $R$ permutation outcomes and calculate the approximate $p$-value.

```{r}
R <- 1000
z <- y - x
n <- length(z)

# generate permutation matrix
outcomes <- replicate(R, sample(c(-1, 1), n, replace = TRUE))

# observed test statistic
zbar <- mean(z)

# test statistic for permutation samples
zl <- apply(outcomes, 2, function(u) mean(u*abs(z)))

# p-value
mean(abs(zl) >= abs(zbar))
```