# Independent $K$-Sample Tests

- $K$ independent samples
- $n_i$: number of observations in group $i$
- $X_{ij}$: $j$th observation from $i$th sample
  - $i \in \{1, \dots, K\}$
  - $j \in \{1, \dots, n_i\}$
- $\mu_i$: mean of group $i$
- One-way ANOVA model: $X_{ij} = \mu_i + e_{ij}$
- Hypotheses: $H_0$: $\mu_1 = \mu_2 = \dots = \mu_K$ vs 
              $H_a$: $H_0$ is not true

For this section, we will use the dataset `iris`.
The column `Sepal.Length` indicates the length of the flower's sepal,
and the column `Species` indicates its species.

```{r}
data("iris")
```


## Classic $F$-Test

- Assumption: $e_{ij} \overset{\mathrm{iid}} \sim N(0, \sigma^2)$
- Mean of group $i$: $\bar{X}_i = \frac{1}{n} \sum_{j = 1}^{n_i} X_{ij}$
- Grand mean: $\bar{X} = \frac{1}{N} \sum_{i=1}^K \sum_{j=1}^{n_i} X_{ij}$
- Treatment sum of squares:
  $$\mathrm{SSTr} = \sum_{i=1}^K n_i (\bar{X}_i - \bar{X})^2$$
- Error sum of squres:
  $$\mathrm{SSE} = \sum_{i=1}^K \sum_{j=1}^{n_i} (X_{ij} - \bar{X}_i)^2$$
- Test statistic:
  $$F_{\mathrm{obs}} = \frac{\mathrm{SSTr}/(K-1)}{\mathrm{SSE}/(N-K)}$$
- Null distribution: Under $H_0$, $F_{\mathrm{obs}} \sim F_{K-1, N-K}$

We can conduct the classic $F$-test using `anova` and `lm`.

```{r}
anova(lm(Sepal.Length ~ Species, data = iris))
```

The SSTr is the value in the first row of column `Sum Sq`.
The SSE is the value in the row `Residuals` of column `Sum Sq`.
The $p$-value is `Pr(>F)`.

## Permutation $F$-Test

- Nonparametric test
- Total number of permutation outcomes: $R = N!/(n_1! n_2! \cdots n_K!)$
- Calculate $F_{\mathrm{obs}}$
- For $\ell \in \{1, \dots, R\}$, 
  $F_\ell$ is the $F$-statistic for the $\ell$th permutation outcome
- We can use the simplified calculation
    $$F \propto \sum_{i=1}^K n_i \bar{X}_i^2$$
- $p\text{-value} = I\{F_\ell \geq F_{\mathrm{obs}}\} / R$

```{r}
# number of permutation samples
R <- 1000 
# number of samples in each group
n <- tapply(iris$Sepal.Length, iris$Species, length)
# total number of observations
N <- sum(n)
# sepal length of each observation
x <- iris$Sepal.Length
# species of each observation
t <- iris$Species

# F-statistic
Fobs <- sum(n * tapply(x, t, mean)^2)

# permutation sample F-statistics
F <- c()
for (i in 1:R) {
  F <- c(F, sum(n * tapply(x, sample(t), mean)^2))
}

# p-value
mean(F >= Fobs)
```


## Kruskal-Walis Test

- Nonparametric test
- Assumption: the data are iid in each group
- Replace data with ranks
- Test statistic: 
  $$\mathrm{KW} = \frac{12}{N(N+1)} \mathrm{SST}^*,$$
  where $\mathrm{SST}^*$ is the SST calculated with ranks
- Null distribution:
  Under $H_0$, $\mathrm{KW} \sim \chi^2_{K-1}$ approximately
  
```{r}
kruskal.test(Sepal.Length ~ Species, data = iris)
```


## Jonckheere-Terpstra Test

- Nonparametric test
- Assumption: the data are iid in each group
- Hypotheses: $H_0$: $\mu_1 = \mu_2 = \dots = \mu_K$ vs 
              $H_a$: $\mu_1 \leq \mu_2 \leq \dots \leq \mu_K$
- Test-Statistic:
  $$\mathrm{JT} = \sum_{i < j} U_{ij},$$
  where $U_{ij}$ is the Mann-Whitney test statistic
  for $H_a$: $\mu_i < \mu_j$
- Approximate null distribution: Under $H_0$,
  $$\frac{\mathrm{JT} - E(\mathrm{JT})}{\sqrt{\mathrm{Var}(\mathrm{JT})}}
  \sim N(0,1)$$
  $$E(\mathrm{JT}) = \frac{N^2 - \sum_{i=1}^K n_i^2}{4}$$
  $$\mathrm{Var}(\mathrm{JT}) 
  = \frac{N^2(2N+3) - \sum_{i=1}^K n_i^2 (2n_i + 3)}{72}$$

```{r}
# assign numbers to each group
species <- factor(iris$Species, levels=c("setosa","versicolor","virginica"))
groups <- as.numeric(species)

sepal <- iris$Sepal.Length

# number of groups
K <- max(groups)

# number of observations
N <-length(sepal)  # total
n <- table(groups) # per group

# compute Uij
JT <- 0
for(i in 1:(K-1)) {
  for(j in (i+1):K) {
     JT <- JT + sum(outer(sepal[groups == i], sepal[groups == j],"<"))
  }
}
JT
```


## Bartlett Test

- Test for equal variances
- Hypotheses: $H_0$: $\sigma_1 = \sigma_2 = \dots = \sigma_K$

```{r}
bartlett.test(iris$Sepal.Length ~ iris$Species)
```


## Multiple Comparisons and Multiple Testing {.tabset .tabset-pills}

We have three methods of deciding the significance levels of multiple comparisons:

Adjustment         Description                             `p.adjust.method`
----------         ---------------------------             -----------------
none               compare $p_i$ with $\alpha$             none
Bonferroni         compare $p_i$ with $\alpha / h$         `"bonferroni"`
Benjamini-Hochberg compare BH-adjusted $p_i$ with $\alpha$ `"BH"`

The Benjamini-Hochberg procedure leads to improved Type I error rate and power.

### Multiple Comparisons

- If we reject $H_0$: $\mu_1 = \mu_2 = \dots = \mu_K$ for a $K$-sample test,
  we can conduct pairwise two-sample tests to determine which pairs of means
  are not equal.
- Total number of pairwise tests for $K$ samples: $h = K(K-1)/2$
- $P(\text{at least one Type I error}) = 1 - (1 - \alpha)^h$

```{r}
pairwise.t.test(iris$Sepal.Length, iris$Species)
```

#### Bonferroni Adjustment

- compare $p$-value for each pairwise comparison $i$ with significance level
  $\alpha_i = \alpha / h$
  - works because $1 - (1 - \alpha_i)^h \approx h \alpha_i$

```{r}
pairwise.t.test(iris$Sepal.Length, iris$Species, p.adjust.method = "bonferroni")
```


### Multiple Testing

- Consider $m$ hypothesis tests together
- The $p$-values are $p_1, \dots, p_h$
- Instead of controlling the Type I error rate $\alpha$, 
  we control the False Discovery Rate
  $$\mathrm{FDR} 
  = E \left( \frac{\text{number of false discoveries}}{\text{number of discoveries}} \right)
  \leq \alpha$$
  - discovery means rejection of $H_0$
- Threshold: Reject $H_0$ if $p\text{-value} \leq t$
- False Discovery Proportion for $i \in \{1, \dots, m\}$

  $$\mathrm{FDP}(t) = \frac{\sum_{i = 1}^m I\{p_i \leq t\} H_i}{\sum_{i = 1}^m I\{p_i \leq t\}}$$
  - $p_i$ is the $p$-value for test $i$
  - $H_i = 1$ if $H_0$ is true and $H_i = 0$ otherwise

#### Benjamini-Hochberg Procedure

- Choose the threshold for rejection to be

  $$t_{\mathrm{BH}} = \max \{P_{(i)} \mid P_{(i)} \leq \frac{\alpha i}{m}, 0 \leq i \leq m\}$$
  - $P_{(1)} \leq P_{(2)} \leq \dots \leq P_{(m)}$
- It can be proved that

  $$\mathrm{FDR} = E[\mathrm{FDP}(t_{\mathrm{BH}})] 
  \leq \frac{\alpha \sum_{i=1}^m H_i}{m}$$

```{r}
pairwise.t.test(iris$Sepal.Length, iris$Species, p.adjust.method = "BH")
```