# One-Sample Tests

This section focuses on hypothesis testing for data in a single sample.

In this section, we will use the weight loss data from class.
Suppose 10 employees are randomly selected to participate
in a weight loss program.
After four weeks, the weight loss of each candidate is given in
`x`.

```{r}
x <- c(-12, 2.1, 0.7, 5, 2.3, 3, 1.9, 1, 1.5, 2.5)
```

We will test the hypothesis $H_a$: $\mu > 0$,
i.e. that the program was effective.

## Parametric Methods

### One-Sample $t$-Test

- Assumption: $n \geq 30$
- Hypotheses: $H_0$: $\mu = \mu_0$ vs $H_a$: $\mu \neq \mu_0$ (or $<$, $>$)
- Test Statistic:
  $$T_{\mathrm{obs}} = \frac{\bar{X} - \mu_0}{s / \sqrt{n}},$$
  - If $\sigma$ is known, ise $\sigma$ instead of $s$
- Null distribution: Under $H_0$, $T_{\mathrm{obs}} \sim t_{n-1}$
  - If $\sigma$ is known, $T_{\mathrm{obs}} \sim N(0,1)$

We use `t.test`  to test $H_a$: $\mu > 0$.

```{r}
t.test(x, alternative = "greater")
```

### Inferences about the Proportion



### Inferences about the Median

## Sign Test

## Wilcoxon Signed-Rank-Test