# Exam Review

## Story A

### Question 1

The exact distribution of $T$ is $\mathrm{Binom}(n, p)$.

### Question 2

$\mathrm{Var}(\hat{p}) = \mathrm{Var}(\frac{1}{n} T) = \frac{1}{n^2} \mathrm{Var}(T) = \frac{1}{n^2}np(1-p) = \frac{1}{n}p(1-p)$

### Question 3

Under large-sample approximation,
   $$Z = \frac{T - np}{\sqrt{np(1-p)}} \sim N(0, 1).$$
```{r}
# significance level
alpha <- 0.05
# critical z-value
1 - qnorm(alpha)
```
   We reject $H_0$ if $Z_{\mathrm{obs}} \geq 1.644854$.
 
### Question 4

Let $n = 10000$ and $T = 15$.
```{r}
T <- 15
n <- 10000
p <- 0.001
# test statistic
zobs <- (T - n * p) / (sqrt(n * p * (1 - p)))
# p-value
1 - pnorm(zobs)
```
    The $p$-value is $0.05673278$.

### Question 5

We can calculate the necessary sample size using the R code below.

```{r}
T <- 15
n <- 10000
p <- 0.0015
for (n in seq(10000, 15000, 1000)) {
  zobs <- (T - n * p) / (sqrt(n * p * (1 - p)))
  pval <- 1 - pnorm(zobs)
  print(c(n, pval))
}
```

The power is $0.9$ somewhere between $n = 13000$ and $n = 15000$,
so we repeat the procedure on that interval.

```{r}
T <- 15
n <- 10000
p <- 0.0015
for (n in seq(13000, 15000, 100)) {
  zobs <- (T - n * p) / (sqrt(n * p * (1 - p)))
  pval <- 1 - pnorm(zobs)
  print(c(n, pval))
}
```
We can see that the power is $0.9$ between $n = 13800$ and $n = 13900$.
Thus, approximately 13,850 impressions are needed to guarantee that the power of the
test is 0.9.

## Story B

```{r}
a <- c(0.01325, 0.00813, 0, 0.01235, 0.00980)
b <- c(0.01613, 0.02532, 0, 0.02020, 0.00503)
```


### Question 6

```{r}
mean(a)
```


The point estimate of $p_A$ is $0.008706$.

### Question 7

Paired is more appropriate.

### Question 8

```{r}
t.test(a, b, paired = TRUE, alternative = "less")$p.value
```

### Question 9

```{r}
wilcox.test(a, b, paired = TRUE, alternative = "less")$p.value
```

### Question 10

idk

### Question 11

The total number of possible permutation outcomes is $R = 2^5 = 32$.

### Question 12

I'm not sure but I think it's (d) grand mean $\bar{X}$.

## Story C

```{r}
clicks <- c(100, 98, 86, 93, 103, 106, 118, 113, 132, 128, 127, 121, 131, 119, 137)
designers <- as.factor(rep(1:3, each = 5))
```

We need the ANOVA table for the subsequent questions.

```{r}
anova(lm(clicks ~ designers))
```


### Question 13

```{r}
mean(clicks)
```

### Question 14

From the ANOVA table, $\mathrm{SSE} = 849.2$.

### Question 15

From the ANOVA table, $\mathrm{SST} = 2610.5$.

### Question 16

From the ANOVA table, $p\text{-value} = 0.0002187$.

### Question 17

```{r}
kruskal.test(clicks, designers)$p.value
```

### Question 18

The total number of permutation outcomes is given by

$$
R = \frac{N!}{n_1! n_2! n_3!} = \frac{15!}{(5!)^3} = 756756
$$
```{r}
factorial(15) / (factorial(5))^3
```

### Question 19

```{r}
# seed
set.seed(10)

# number of samples in each group
n <- c(5, 5, 5)

# F-statistic
Fobs <- sum(n * tapply(clicks, designers, mean)^2)

# number of permutation samples
R <- 1000 
# empty vector for the F-statistic of each permutation sample
F <- c()
# calculate permutation sample F-statistics
for (i in 1:R) {
  F <- c(F, sum(n * tapply(clicks, sample(designers), mean)^2))
}

# p-value = proportion of permutation F-statistics greater than Fobs
mean(F >= Fobs)
```

### Question 20

Jonckheere-Terpstra Test